Summary Statistics

### 2025-03-13 extensive_tests/draft-sumstat-vignette.Rmd
Updating vignette with sumstat for mean number of significat freq changes per island and cherry

>>>>

### Mean Number of Significant Methylation Frequency Changes per Island and Cherry

A cherry $c$ is defined as a pair of direct offspring (tips \( t_1 \) and \( t_2 \)) 
of the same internal node in the phylogenetic tree. 
For a given cherry $c$, let $\mathcal{I}$ be the set of indices for island structures. 
For each island structure $i \in \mathcal{I}$ at each cherry tip $t_c \in {t_{c,1}, t_{c,2}$ in each phylogenetic tree cherries, 
the number of sites in each 
methylation state (unmethylated \( 0 \), partially-methylated \( 0.5 \), 
methylated \( 1 \)) is counted. This count is represented as:

$$
\text{count}_{\text{UPM}}(c,t_c, i) = \left( n_0, n_{0.5}, n_1 \right)
$$

where \( n_0\), \( n_{0.5}\), and \( n_1\) are the counts of unmethylated, 
partially-methylated, and methylated sites at tip \( t_c\), respectively.

The distribution of methylation states at each island between the two tips \( t_1 \) 
and \( t_2 \) of a given cherry is compared using a chi-squared test. 
The null hypothesis \( H_0 \) is that the distributions of methylation states at 
the two tips are the same. The test statistic is calculated using the contingency 
table \( T \) of the methylation state counts for each tip:

$$
H_0: \text{distribution}(t_{c,1}) = \text{distribution}(t_{c,2})
$$

The contingency table \( T_{c,i} \) for tips \( t_{c,1} \) and \( t_{c,2} \) is:

$$
T_{c,i} = \begin{pmatrix}
n_0^{t_{c,1}} & n_{0.5}^{t_{c,1}} & n_1^{t_{c,1}} \\
n_0^{t_{c,2}} & n_{0.5}^{t_{c,2}} & n_1^{t_{c,2}}
\end{pmatrix}
$$

The p-value for the chi-squared test is calculated via Monte Carlo simulation to 
improve reliability, even when the expected frequencies do not meet the assumptions 
of the chi-squared test (i.e., expected counts of at least 5 in each category).

For each cherry $c$ and each island $i$, the methylation data at each cherry tip 
for the specified genomic structures (islands) are compared using the chi-squared test.

$$
p_{\text{value}}(c,i) = \chi^2\text{-test}(T_{c,i})
$$

These p-values are stored in a dataframe for each cherry and for each island.

The significance of the methylation frequency changes is determined based on a 
user-defined threshold \( p_{\text{threshold}} \). If the p-value is smaller than 
the threshold, the change is considered significant. This is done for each cherry 
and each island \( i \in \mathcal{I} \) by comparing each p-value to the threshold:

$$
\text{Significant Change}_{c,i} = 
\begin{cases} 
1, & \text{if } p_{\text{value}}(c,i) < p_{\text{threshold}} \\
0, & \text{if } p_{\text{value}}(c,i) \ge p_{\text{threshold}}
\end{cases}
$$

For each cherry , the mean number of significant changes across all islands 
\( i \in \mathcal{I} \) is computed as the average of the significant changes for each island:

$$
\text{Mean Number of Significant Changes per Island}_{c} = \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} \text{Significant Change}_{c,i}
$$

where \( |\mathcal{I}| \) is the total number of islands.
